---
title: "Opioid use in Cancer Patients"
author: "Matthew Siddoway"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

# BACKGROUND

# STUDY QUESTION and HYPOTHESIS

## Questions  
What effect does genetics and phenotypes have on pain management among European cancer patients who have been treated with opioids?

## Hypothesis 
Genetic variation in opioid receptor-related genes significantly affect pain relief and opioid requirements.
## Prediction
I predict that there will be certain phenotypes and genetics that correspond to lesser pain responses depending on the unique genetic makeup of each demographic.

# METHODS
```{r}

library(ggplot2)
data <- read.csv("Opioid dataset.csv")
str(data)

model <- lm(percent_pain_relief ~ daily_dose_mg_ME + age + sex +
               rs12211463_minor_allele_count +
               rs13421094_minor_allele_count +
               rs10413396_minor_allele_count,
             data = data)

summary(model)


ggplot(data, aes(x = daily_dose_mg_ME, y = percent_pain_relief, color = sex)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Relationship Between Opioid Dose and Pain Relief",
    subtitle = "Simulated data based on Galvan et al. (2011)",
    x = "Daily Dose (Morphine Equivalent, mg)",
    y = "Percent Pain Relief"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```
## Barplots


## Generalized Linear Model


# DISCUSSION


## Interpretation of barplots


## Interpretation of the generalized linear model


# CONCLUSION


# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.



